---
output: html_document
editor_options: 
  chunk_output_type: console
---
pu--
title: "HW2"
format: html
---

In the 1930s, the Home Owners’ Loan Corporation (HOLC) categorized neighborhoods according to an arbitrary estimate of safety relevant to real estate investing. Categories ranged from A (green), B (blue), C (yellow), D (red). This meant homes in the red category would rarely get home loans, over time creating wealth and health. This phenomenon is known as "redlining". Recently, there was a study finding negative effects of redlining on biodiversity observations in redlined communities. 

In this project, we will investigate environmental (in)justice metrics and biodiversity observations in correlation with HOLC grades in Los Angeles, California, United States.

```{r}
# read in packages
library(here)
library(tmap)
library(sf)
library(tidyverse)
```

Part 1: Explore historical redlining in Los Angeles and its legacy on present-day environmental justice

1. Create a map of historical redlining neighborhoods, including:
- neighborhoods colored by HOLC grade
- an appropriate base map
- a clear title and legend

```{r}
# read in HOLC data
holc <- st_read(here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))
unique(holc$fill)

ej_screen <- st_read(here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) %>% 
  select(ID, STATE_NAME, ST_ABBREV, CNTY_NAME, REGION, ACSTOTPOP, LOWINCPCT, P_PM25, P_LIFEEXPPCT, Shape_Length, Shape_Area, Shape)
```

```{r}
# map of neighborhood grades
tm_shape(holc) +
  tm_polygons(fill = "fill") +
  tm_basemap("CartoDB.PositronNoLabels") +
  tm_add_legend(labels = c("Grade A", "Grade B", "Grade C", "Grade D"),
                fill = c("#76a865","#7cb5bd","#ffff00","#d9838d"),
                title = "HOLC Grade Level",
                position = tm_pos_in("right", "bottom")) +
  tm_title("HOLC Neighborhood Grades in Los Angeles") +
  tm_scalebar(position = tm_pos_in("right", "top")) +
  tm_compass(position = tm_pos_in("left", "bottom"))
```

```{r}
# summary table 

# combine ej screen data and holc data

# check if crs match
st_crs(ej_screen) == st_crs(holc) # FALSE
ej_screen_transform <- st_transform(ej_screen, crs = st_crs(holc))
st_crs(ej_screen_transform) == st_crs(holc) # now true

# check if both are sf objects
class(holc)
class(ej_screen)

st_crs(ej_screen_transform)
st_crs(holc)

# join 
ej_holc <- st_join(holc, ej_screen_transform, left = TRUE)


```

